---
title: "Essay"
author: '11484265'
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(HMDHFDplus)
library(viridis)
library(rstan)

source('overview.R')
source('models.R')
```


# Overview (10%)
-mort patterns by age and over time in both sexes (2 graphs total)
```{r setup}

mort_over_f
mort_over_m

```



# Forecasts (50%)
-split into train and forecast data
--75% would be until ~1998, ok for long term, but may not capture recent shifts, so decided 2010
--opiod crisis in 2016
--covid
--babyboomer first wave retire 2011
--choose 2012 (leave 10y, 2013-2022 of test data), 

## Log-Linear Model
- forecast results

###warnings
Warning messages:
1: There were 1 chains where the estimated Bayesian Fraction of Missing Information was low. See
https://mc-stan.org/misc/warnings.html#bfmi-low 

FIXED BY INCREASING adapt_delta FROM 0.97 TO 0.9999 and warm up to 2000

for male all 4 chains have low BFMI but rhat is close to 1, so convergence is still good.
No divergence warnings.
Trace plots should show stationary, non-trending behavior.
Density plots are smooth and unimodal (unless there are multiple modes in the posterior).

2: Examine the pairs() plot to diagnose sampling problems
 
3: The largest R-hat is 1.06, indicating chains have not mixed.
Running the chains for more iterations may help. See
https://mc-stan.org/misc/warnings.html#r-hat 

FIXED BY INCREASING ITER FROM 1500 TO 5000


4: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.
Running the chains for more iterations may help. See
https://mc-stan.org/misc/warnings.html#bulk-ess 

FIXED BY INCREASING ITER FROM 1500 TO 6000
increased chains to 4


5: Tail Effective Samples Size (ESS) is too low, indicating posterior variances and tail quantiles may be unreliable.
Running the chains for more iterations may help. See
https://mc-stan.org/misc/warnings.html#tail-ess

FIXED BY INCREASING ITER FROM 1500 TO 5000

## Lee-Carter Model
- forecast results

### warnings
for BOTH SEXES 
check that the rhat is only for ben (check traceplot converges for each parameter), and look at rhat values (how???)

Warning messages:
1: The largest R-hat is 1.53, indicating chains have not mixed.
Running the chains for more iterations may help. See
https://mc-stan.org/misc/warnings.html#r-hat 
2: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.
Running the chains for more iterations may help. See
https://mc-stan.org/misc/warnings.html#bulk-ess 
3: Tail Effective Samples Size (ESS) is too low, indicating posterior variances and tail quantiles may be unreliable.
Running the chains for more iterations may help. See
https://mc-stan.org/misc/warnings.html#tail-ess 


- show results

## Quality of Estimates
- graphs of convergence of each parameter 
- graph of rhat?
- forecasts of all age groups compared to actual (lc and ln, so 4 graphs total)
```{r}
gg_mort_f
gg_mort_m
```

- forecastts of life expectancy (age v year, median, lc and ln, 4 graphs total)

```{r}

```



#### compare quality (numbers)

Using mean of kf (posterior mort) since it's fiarly symmetrical/normal

```{r}

tab_me
tab_mpe
tab_mae
tab_rmse
tab_mape

```


using the 95% credible interval (columns 2.5% and 97.5% ), calculate the proportion of observed
mortality rates in 2010-2019 that fall within the predictive intervals of the forecasts. Compare these
measures between log-linear and Lee-Carter models.
```{r}
prop_within_ln_f
prop_within_lc_f
prop_within_ln_m
prop_within_lc_m
```



# Challenges (40%)
## short-term (5y)
## long-term (30-50y)