---
title: "Essay"
author: '11484265'
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(HMDHFDplus)
library(viridis)
library(rstan)

source('overview.R')
source('models.R')
```


# Overview 

The graphs show mortality trends for males and females in Canada from 1921 to 2022, with mortality rates measured in log10 scale. For both sexes, mortality is highest in infancy and early childhood, as indicated by the steep dip at age 0 and then a gradual increase starting from early adulthood. In males, mortality consistently increases with age, with a noticeable uptick in middle adulthood, particularly post-50. The mortality rate for males also shows a marked upward trend after 60, peaking around 100 years old. For females, the trend follows a similar pattern, but the rate of increase with age is slightly less pronounced compared to males, especially in younger and middle adulthood.

```{r setup}

mort_over_f
mort_over_m

```

Over time, mortality for both sexes has decreased significantly, particularly from the 1920s through to the early 2000s. The data shows a distinct shift from darker purple hues (indicating higher mortality) in earlier years to lighter yellow and orange hues (indicating lower mortality) in the later years, reflecting improvements in healthcare and living conditions. The trends suggest a steady improvement in life expectancy, with sharper declines in mortality observed after the mid-20th century, though the highest mortality rates still occur in the older age groups. Notably, female mortality rates remain slightly lower than those of males across the entire age range, suggesting a consistent gender gap in survival.

# Forecasts 

I'm using 1921 to 2012 as training data, and 2013-2022 as test. 2011-13 is roughly when the first wave of babyboomer retires. 75% would be until ~1998, ok for long term, but may not capture recent shifts, so decided 2013.

forecasts of all age groups compared to actual
```{r gg_fore_mort}
gg_mort_f
gg_mort_m
```


## Log-Linear Model

- iSSUE:  Bayesian Fraction of Missing Information was low, increase adapt_delta
FIXED BY INCREASING adapt_delta FROM 0.97 TO 0.9999 and warm up to 2000

- iSSUE:  The largest R-hat is 1.06, Bulk Effective Samples Size (ESS) AND tail Effective Samples Sizeis too low, indicating posterior means and medians may be unreliable.  
FIXED BY INCREASING ITER FROM 1500 TO 6000
increased chains to 4

for male all 4 chains have low BFMI but rhat is close to 1, so convergence is still good.
No divergence warnings.
SHOW GRAPH?????
- Trace plots should show stationary, non-trending behavior.
- Density plots are smooth and unimodal (unless there are multiple modes in the posterior).


## Lee-Carter Model
- forecast results

### warnings
for BOTH SEXES 
-issue: large r-hat(1.53), chains have not mixed (only ben?) 
```{r}
lc_trace_ben_f 
lc_trace_ben_m
```

-issue: bulk and tail ESS too low, indicating posterior means and medians may be unreliable.

## Quality of Estimates
- graphs of convergence of each parameter 
- graph of rhat?

- forecastts of life expectancy (age v year, median, lc and ln, 4 graphs total)

```{r gg_fore_le}
gg_le_f
gg_le_m
```



#### compare quality (numbers)

Using mean of kf (posterior mort) since it's fiarly symmetrical/normal

```{r}

tab_me
tab_mpe
tab_mae
tab_rmse
tab_mape

```


using the 95% credible interval (columns 2.5% and 97.5% ), calculate the proportion of observed
mortality rates in 2010-2019 that fall within the predictive intervals of the forecasts. Compare these
measures between log-linear and Lee-Carter models.

```{r}
prop_within_ln_f
prop_within_lc_f
prop_within_ln_m
prop_within_lc_m
```

# Challenges (40%)
## short-term (5y)
## long-term (30-50y)